A New Method of Modelling Tuneable Lasers with Functional Composition

### 2019-07-12 Thesis/Model/Model.tex
Finished final read through.

\subsection{Loss}
Two sources of loss exist within the laser circuit: the loss due to the output coupler and the optical loss due to absorption and scattering. The first case is simply a scalar multiplication depending on the reflectivity of the output coupler. The loss due to absorption and scattering can be derived from \eqref{eq:nlse},
\begin{align*}
\pdiff{A}{z} = - \alpha A.
\end{align*}
Combining these two effects gives a loss that takes the form
\begin{align*}
L(A) &= (1 - R) \textrm{e}^{- \alpha L_T}A,
\end{align*}
where $R$ is the reflectivity of the output coupler, and $L_T$ is the total length of the laser circuit.

\subsection{Dispersion}
Within the laser cavity, the dispersion is dominated by the \gls{cfbg}. In comparison, the dispersion due to the fibre is negligible. The dispersive terms of \eqref{eq:nlse} give
\begin{align}
\label{eq:dispde}
	\pdiff{A}{z} = -i \frac{\beta_2}{2} \pdiff[2]{A}{T},
\end{align}
and since dispersion acts in the frequency domain it is convenient to take the Fourier transform\footnote{We use $\displaystyle \FT{f} = (2 \pi)^{1/2} \int_\mathbb{R} f(x) \textrm{e}^{i \omega x} \df x$ as the definition of the Fourier transform.} of \eqref{eq:dispde} \cite{debnath, gradshteyn}, giving
\begin{align*}
	\pdiff{}{z}\FT{A} &= i\frac{\omega^2}{2}\beta_2 \FT{A}.
\end{align*}
The effect of dispersion is then by the map
\begin{align*}
D(A) &= \FTi{\textrm{e}^{i \omega^2 L_D\beta_2/2} \FT{A}},
\end{align*}
where $L_D$ is the length of the dispersive medium. \\

@@

The structure of each process of the laser can be better understood by re-scaling the time, energy, and amplitude. Nominal values for tuneable lasers are shown in Table~\ref{tab:values}. Knowing experimental durations and energies, the table suggests the convenient scalings:
\begin{align}
	T &= T_M \widetilde{T},& E &= \Es \widetilde{E},& A &= \left( \frac{\Es}{T_M} \right)^{1/2} \widetilde{A}.
\end{align}
Revisiting each process map shows each process has a characteristic non-dimensional parameter. The new mappings, after dropping the tildes, are
\begin{subequations}
\label{eq:effects}
\begin{align}
\label{eq:gain}
G(A) &= \left(E^{-1} W \left( a E \textrm{e}^{E}\right) \right)^{1/2} A, \\
\label{eq:fibre}
F(A) &= A \textrm{e}^{i b |A|^2}, \\
\label{eq:loss}
L(A) &= h A, \\
\label{eq:disp}
D(A) &= \FTi{\textrm{e}^{i s^2 \omega^2} \FT{A}}, \\
\label{eq:mod}
M(A) &= \textrm{e}^{-T^2 / 2} A,
\end{align}
\end{subequations}
with the four dimensionless parameters, as defined by the values in Table \ref{tab:values},
\begin{equation}
\begin{aligned}
	a &= \textrm{e}^{g_0 L_g} \sim 8 \times 10^3,& \qquad h &= (1 - R) \textrm{e}^{-\alpha L} \sim 0.04\\
	b &= \gamma L_f \frac{\Es}{T_M} \sim 1,& \qquad s &= \sqrt{\frac{\beta_2 L_D}{2 T_M^2}} \sim 0.2,
\label{eq:ndparam}
\end{aligned}
\end{equation}
which characterize the behaviour of the laser. Notice that the modulation is characterized by $T_M$, and each other process has it's own independent non-dimensional parameter.

@@

\section{Solution to the Linear Model}
\label{sec:linear}
In reality the nonlinearity parameter, $b$, is of order unity by \eqref{eq:ndparam}, but we shall start our investigation with the simpler case of $b = 0$---which shall be referred to as the linear model. In this case, a solution can be found analytically. It is expected the solution will take the form of a Gaussian. There are a few reasons for this; the solution to the previous models were Gaussian \cite{cutler, siegman, kuizenga1970a, martinez1984, martinez1985} (see Section \ref{chap:meml}), the equilibrium shape will be highly correlated to the shape of the modulation function, and since a Gaussian is a fixed point of the Fourier transform \cite{gradshteyn}, dispersion will not alter the envelope. \\

@@

\subsection{Asymptotic Expansion of the Variance}
\begin{figure}[tbp]
\centering
\input{./Figures/Asympt}
\caption{Plot of \eqref{eq:asymp} and \eqref{eq:asymp2} to highlight the structure.}
\label{fig:asympt}
\end{figure}
While \eqref{eq:equilvar} is useful, so too is the asymptotic behaviour. The asymptotic expansion provides additional information about the structure that the full analytic expression cannot. Moveover, the expansion provides more detail about the limits, as well has the rate of convergence. Lastly, the expansion allows for simplified expressions within their domains that are much less unwieldy than those found in the previous subsections. \\

@@

We can now find the amplitude of the pulse as well now that the energy is know---as we have previously found the equilibrium shape. Again, from \eqref{eq:gainde}, it must be that $E_* = \sqrt{\pi} P \sigma$, or,
\begin{align}
P = \frac{W(a E \textrm{e}^E)}{\sqrt{\pi} \sigma}.
\label{eq:equilpower}
\end{align}
The asymptotic expansions when $s \rightarrow 0$ of the energy, and peak power are---for completeness---
\begin{align}
E_* &= \Lambda - \Theta \frac{\Lambda}{1 + \Lambda} s + \bigO{s^2}, & P &= \frac{\Lambda}{\sqrt{2 \pi s}} + \left( \frac{\Lambda}{2} -  \Theta \frac{\Lambda}{1 + \Lambda} \right) \sqrt{\frac{s}{2 \pi}} +  \bigO{s^{3/2}},
\label{eq:epasymp}
\end{align}
respectively, where
\begin{align*}
\Lambda &= W\left( \frac{a h^2 \ln(a h^2)}{1 - h^2} \exp \left( \frac{h^2 \ln(a h^2)}{1 - h^2} \right) \right),
\end{align*}
and
\begin{align*}
\Theta &= \frac{\left( h^2 \ln(a h^2) + 1 - h^2 \right) \left( \ln(a h^2) + 1 - h^2 \right)}{\left( 1 - h^2 \right)^2 \ln(a h^2)}.
\end{align*}

\section{Chapter Summary}
In this chapter we first derived the effect each of the five components---gain, nonlinearity, loss, dispersion, and modulation---has on the incoming pulse from \eqref{eq:nlse}. Furthermore, to highlight the structure of these effects the system was non-dimensionalized. This yielded the functional maps \eqref{eq:effects} for the components, with corresponding non-dimensional parameters \eqref{eq:ndparam}. With each of the operators defined, we proceeded to compose each process, in the order most representative of a tuneable laser, to give the effect of one complete circuit around the laser cavity. \\

>>>>

\subsection{Loss}
Two sources of loss exist within the laser circuit: the loss due to the output coupler and the optical loss due to absorption and scattering. The first case is simply a scalar multiplication depending on the reflectivity of the output coupler. The loss due to absorption and scattering can be derived from \eqref{eq:nlse},
\begin{align*}
\pdiff{A}{z} = - \alpha A,
\end{align*}
and gives $\textrm{e}^{-\alpha L_T}A$, which by multiplying by the amount of the signal lost to the output coupler gives
\begin{align*}
L(A) &= (1 - R) \textrm{e}^{- \alpha L_T}A,
\end{align*}
where $R$ is the reflectivity of the output coupler, and $L_T$ is the total length of the laser circuit as the effect of the losses\footnote{Depending on the experimental setup the loss may take the form $L(A) = R \textrm{e}^{- \alpha L_T}A$.}.

\subsection{Dispersion}
Within the laser cavity, the dispersion is dominated by the \gls{cfbg}. In comparison, the dispersion due to the fibre is negligible. The dispersive terms of \eqref{eq:nlse} give
\begin{align}
\label{eq:dispde}
	\pdiff{A}{z} = -i \frac{\beta_2}{2} \pdiff[2]{A}{T},
\end{align}
and since dispersion acts in the frequency domain it is convenient to take the Fourier transform\footnote{We use $\displaystyle \FT{f} = (2 \pi)^{1/2} \int_\mathbb{R} f(x) \textrm{e}^{i \omega x} \df x$ as the definition of the Fourier transform.} of \eqref{eq:dispde} \cite{debnath, gradshteyn}, giving
\begin{align*}
	\pdiff{}{z}\FT{A} &= i\frac{\omega^2}{2}\beta_2 \FT{A}.
\end{align*}
The effect of dispersion is then given by the map
\begin{align*}
D(A) &= \FTi{\textrm{e}^{i \omega^2 L_D\beta_2/2} \FT{A}},
\end{align*}
where $L_D$ is the length of the dispersive medium. \\

@@

The structure of each process of the laser can be better understood by re-scaling the time, energy, and amplitude. Nominal values for tuneable lasers are shown in Table~\ref{tab:values}. Knowing experimental durations and energies, the table suggests the convenient scalings:
\begin{align}
	T &= T_M \widetilde{T},& E &= \Es \widetilde{E},& A &= \left( \frac{\Es}{T_M} \right)^{1/2} \widetilde{A}.
\end{align}
Revisiting each process map shows each process has a characteristic non-dimensional parameter. The new mappings, after dropping the tildes, are
\begin{subequations}
\label{eq:effects}
\begin{align}
\label{eq:gain}
G(A) &= \left(E^{-1} W \left( a E \textrm{e}^{E}\right) \right)^{1/2} A, \\
\label{eq:fibre}
F(A) &= A \textrm{e}^{i b |A|^2}, \\
\label{eq:loss}
L(A) &= h A, \\
\label{eq:disp}
D(A) &= \FTi{\textrm{e}^{i s^2 \omega^2} \FT{A}}, \\
\label{eq:mod}
M(A) &= \textrm{e}^{-T^2 / 2} A,
\end{align}
\end{subequations}
with the four dimensionless parameters, as defined by the values in Table \ref{tab:values},
\begin{equation}
\begin{aligned}
	a &= \textrm{e}^{g_0 L_g} \sim 8 \times 10^3,& \qquad h &= (1 - R) \textrm{e}^{-\alpha L} \sim 0.04, \\
	b &= \gamma L_f \frac{\Es}{T_M} \sim 1,& \qquad s &= \sqrt{\frac{\beta_2 L_D}{2 T_M^2}} \sim 0.2,
\label{eq:ndparam}
\end{aligned}
\end{equation}
which characterize the behaviour of the laser. Notice that the modulation is characterized by $T_M$, and each other process has its own independent non-dimensional parameter.

@@

\section{Solution to the Linear Model}
\label{sec:linear}
In reality the nonlinearity parameter, $b$, is of order unity by \eqref{eq:ndparam}, but we shall start our investigation with the simpler case of $b = 0$---which shall be referred to as the linear model. In this case, a solution can be found analytically. It is expected the solution will take the form of a Gaussian. There are a few reasons for this; the solution to the previous models were Gaussian \cite{cutler, siegman, kuizenga1970a, martinez1984, martinez1985} (see Section \ref{chap:meml}), the equilibrium shape will be highly correlated to the shape of the modulation function, and since a Gaussian is a fixed point of the Fourier transform \cite{gradshteyn} dispersion will not alter the envelope. \\

@@

\subsection{Asymptotic Expansion of the Variance}
\begin{figure}[tbp]
\centering
\input{./Figures/Asympt}
\caption{Plot of \eqref{eq:asymp} and \eqref{eq:asymp2} to highlight the structure.}
\label{fig:asympt}
\end{figure}
While \eqref{eq:equilvar} is useful, so too is the asymptotic behaviour. The asymptotic expansion provides additional information about the structure that the full analytic expression cannot. Moveover, the expansion provides more detail about the limits, as well has the rate of convergence. Lastly, the expansion allows for simplified expressions within their domains that are much less unwieldy than those found in the previous subsection. \\

@@

We can now find the amplitude of the pulse as well now that the energy is known---as we have previously found the equilibrium shape. Again, from \eqref{eq:gainde}, it must be that $E_* = \sqrt{\pi} P \sigma$, or,
\begin{align}
P = \frac{W(a E \textrm{e}^E)}{\sqrt{\pi} \sigma}.
\label{eq:equilpower}
\end{align}
The asymptotic expansions when $s \rightarrow 0$ of the energy, and peak power are---for completeness---
\begin{align}
E_* &= \Lambda - \Theta \frac{\Lambda}{1 + \Lambda} s + \bigO{s^2}, & P &= \frac{\Lambda}{\sqrt{2 \pi s}} + \left( \frac{\Lambda}{2} -  \Theta \frac{\Lambda}{1 + \Lambda} \right) \sqrt{\frac{s}{2 \pi}} +  \bigO{s^{3/2}},
\label{eq:epasymp}
\end{align}
respectively, where
\begin{align*}
\Lambda &= W\left( \frac{a h^2 \ln(a h^2)}{1 - h^2} \exp \left( \frac{h^2 \ln(a h^2)}{1 - h^2} \right) \right),
\end{align*}
and
\begin{align*}
\Theta &= \frac{\left( h^2 \ln(a h^2) + 1 - h^2 \right) \left( \ln(a h^2) + 1 - h^2 \right)}{\left( 1 - h^2 \right)^2 \ln(a h^2)}.
\end{align*}

\section{Chapter Summary}
In this chapter we first derived the effect each of the five components---gain, nonlinearity, loss, dispersion, and modulation---has on the incoming pulse from \eqref{eq:nlse}. Furthermore, to highlight the structure of these effects the system was non-dimensionalized. This yielded the functional maps \eqref{eq:effects} for the components, with corresponding non-dimensional parameters \eqref{eq:ndparam}. With all of the operators defined, we proceeded to compose each process, in the order most representative of a tuneable laser, to give the effect of one complete circuit around the laser cavity. \\

### 2019-07-12 Thesis/Nonlinear/Nonlinear.tex
Finished final read through.

\subsection{Validation}
\begin{figure}[tbp]
\centering
\input{./Figures/Variance}
\caption[Equilibrium variance, chirp, and phase shift of the pulse as a function of the dispersion parameter.]{Simulation and analytic equilibrium variance, chirp, and phase shift as a function of the dispersion parameter, $s$. The analytic expressions are given by \eqref{eq:equilvar}, \eqref{eq:equilchirp}, \eqref{eq:phase}, respectively.}
\label{fig:var}
\end{figure}
Before delving into the nonlinear behaviour of the model, we wish to validate that the code is working as expected by comparing the results of the simulations with $b=0$ to the results of the linear model in Chapter \ref{chap:linear}. In the case of the numerical solution, our initial conditions become somewhat important for two reasons. First, if the energy is too high initially the nonlinear effects are too strong leading to \acrlong{spm}. Secondly, the initial waveform needs to be well behaved enough so that the Fourier transform is band limited. In all of the following analysis the initial waveform is $\Gamma \sech \left( 2T \right) \textrm{e}^{i \pi / 4}$ normalized so that the initial energy is $E_0 = 0.1$; additionally, $a = 8000$, and $h = 0.04$. \\

@@

In the first experiment the pulse is allowed to equilibrate for 40 loops of the circuit\footnote{The choice of 40 will be justified in the following section.}, we then compare the variance, chirp, and phase shift of the analytic and numerical solution. The results of this are shown in Figure \ref{fig:var}. For the most part, we see exceptional agreement between the analytic solution and numerical solution for the variance, and phase shift. However, as $s \rightarrow 2.5$ the chirp from the simulations begins to deviate from the analytic solution. The reason for this is quite a simple one. We shall turn our attention to the energy, and amplitude of the pulse at equilibrium as shown in Figure \ref{fig:valenergy}. As with Figure \ref{fig:var}, there is very good agreement between the two solutions. It is also now clear that at approximately $s = 2.5$ there is too much dispersion---the pulse is no longer sustainable. This is of course a consequence of the condition \eqref{eq:energycond}. Since the pulse effectively vanishes after this point, the chirp calculation becomes numerically unstable leading to the disagreement. \\

With the numerical solution yielding the expected results we are now ready to explore the rich structure, which the nonlinearity adds.

\section{Nonlinear Model}
\begin{figure}[tbp]
\centering
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Shape}
\caption{Envelope}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_FT}
\caption{Fourier transform}
\end{subfigure} \\
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Chirp}
\caption{Chirp}
\end{subfigure}
\caption[Envelope, Fourier transform, and chirp of the pulse---stable case.]{Simulation with $s = 0.15$, and $b = 2.1$ after 15 circuits.}
\label{fig:nlstable}
\end{figure}
With the inclusion of the nonlinearity we generally find a similar solution to the linear case. An example of this is shown in Figure \ref{fig:nlstable}. The envelope of the pulse is unsurprisingly Gaussian-esque, however, it is \emph{not} precisely Gaussian and more closely resembles a generalized Gaussian\footnote{A generalized Gaussian has the form $\exp \left(-t^\alpha \right)$, with $\alpha > 2$.}. The fact that the pulse envelope is not Gaussian is further emphasized while examining the Fourier transform of the pulse. If the pulse were a Gaussian, we would expect the Fourier transform to also be a Gaussian \cite{debnath, gradshteyn}. Instead, the magnitude of the Fourier transform has a unique weakly bi-modal shape. This deviation suggests the nonlinearity implants higher frequency oscillations into the pulse---a key observation in the coming subsections. Finally, we shall examine the derivative of the phase---essentially the chirp. Recall in Chapter \ref{chap:linear} the chirp was defined as coefficient of $-\frac{1}{2}iT^2$ in the exponential, and by taking the negative of the derivative we would expect a linear function with a slope equal to the chirp. In the nonlinear case, we recover this linear response for moderate values of $T$. However, for $|T| > 1$ the chirp begins to saturate. This is consistent with experimental results \cite{chen, rothenberg, tomlinson}. \\

@@

\begin{figure}[tbp]
\centering
\input{./Figures/ROC}
\caption{Rate of convergence as a function of number of iterations for various $s$ values, and $b = 0.5$. The error is given by \eqref{eq:error}.}
\label{fig:roc}
\end{figure}

@@

The energy of the pulse is directly related to the output power of the laser, and since this is not as controllable as in a regular laser, it is of great interest. In this nonlinear case we allow the initial pulse to pass through each component 100 times instead of 40 in the linearized case. This choice can be justified by Figure \ref{fig:roc}. The pulse reaches a state in which the error no longer decreases within typically 10--20 iterations. In the linearized case we chose to iterate 40 times since this is about double the number of iterations expected. Furthermore, in the nonlinear case we extend this number further to 100 in case the nonlinearity increases the number of iterations to converge. The energy of the pulse after 100 iterations is shown in Figure \ref{fig:energy}. In Figure \ref{fig:energyzoom} we generally find that the energy is a smooth surface that slowly decays as $s$, and $b$ increase. In the upper left hand region of the plot this is clearly not the case. The contours show that the energy is very noisy with very large derivatives. Perhaps more surprising, as shown in Figure \ref{fig:energybig}, is that this boundary between periodic and aperiodic behaviour appears to become periodic (in the parameter space) as $b$ increases around $s = 0.3$. These features are further highlighted in Figure \ref{fig:cartoon} by disregarding the energy and only examining whether the pulse is invariant after additional iterations. \\

@@

The noise in the energy exhibited for moderate to large values of $b$, and small values of $s$ is a phenomenon called \emph{wave breaking} \cite{agrawal2013, anderson, finot, rothenberg, tomlinson}. Wave breaking is not limited to just optics; wave breaking occurs in areas such as plasmas, transmission lines, and fluid dynamics \cite{rothenberg}. Wave breaking occurs because the pulse begins to interfere with itself in a way called \gls{spm} \cite{agrawal2002, agrawal2013, becker}. \gls{spm} occurs because the index of refraction is intensity dependent \cite{agrawal2002, becker, rothenberg, silfvast}, which leads to additional chirp across the pulse \cite{agrawal2013, anderson, rothenberg, silfvast}. This in turn causes higher order frequencies to be injected into the pulse \cite{agrawal2013, anderson}, as we saw in Figure \ref{fig:nlstable}. These high frequencies compound with each trip around the cavity becoming parasitic very quickly---Figure \ref{fig:break} highlights this. Notice that the difference between Figure \ref{fig:nlstable} and Figure \ref{fig:break} is a difference in $b$ of $0.05$---just enough to cross the boundary---this difference could be attributed to adding a few centimetres more of fibre between the gain and output coupler. The left figures show the pulse after 11 trips around the cavity; in the Fourier transform it is clear that the contributions from higher frequencies has increased. We obtain comparable results as in the experiments \cite{anderson, rothenberg}. Additionally, the chirp starts losing its linearity causing it to start becoming unstable; the nature of this instability is again in agreement with the experiments \cite{anderson, rothenberg}. The parasitic nature of the high frequency contributions is evident by examining the right figures. After five additional trips around the cavity, the envelope of the pulse is much more rippled, and the real and imaginary parts become incoherent. Moreover, the Fourier transform has no clear structure and has essentially become noise. The chirp has grown to be highly oscillatory and unstable. Once the pulse has reached a state such as this---the envelope, Fourier transform, and chirp never reach a steady equilibrium state. \\

@@

Unsurprisingly, the error is largest in the region where the wave breaks. As mentioned in the previous subsection, the pulse does not reach a stable state in this region. As a consequence the envelope varies drastically, which leads to this large error. Conversely, the region where the energy appeared to be stable still has a non-zero error. The cause of this discrepancy is because some equilibrium states have a periodicity greater than $1$, thus, we redefine \eqref{eq:error} to be instead
\begin{align}
\textrm{E}_\Delta = \frac{\| |A_i| - |A_{i-\Delta}| \|_2}{\| A_{i-\Delta} \|_2},
\label{eq:deltaerror}
\end{align}
to pick up equilibrium states with period $\Delta$, again with $i = 100$. Figure \ref{fig:deltaerror} shows the error for an assortment of $\Delta$ values. There are several key observations to make, first, for $\Delta = 2$ the kidney-shaped blob in the lower right corner has vanished, thus, this entire region has a periodicity of $2$. Second, for $\Delta = 3$ in the upper right corner a negative triangular-shaped structure emerges. Lastly, $\Delta \in \left\{ 5, 7, 11, 13 \right\}$ may not seem to extract any higher order periods, however, there are in fact very small striations in the lower left. We typically chose $\Delta$ values to be prime since they are the building blocks for composite numbers. However, there are two composite $\Delta$ values of interest. The first is $16$, since $16 = 2^4$ and it will pick up any periodic behaviour with orders of $2, 4, 8, 16$. Lastly, we chose $\Delta = 60$ for a similar reason: because $60$ is highly divisible extracting orders such as $12$ or $30$ without having to explicitly run the calculations\footnote{In the case of $\Delta = 60$ a value of $i = 150$ is used to ensure sufficient iterations to converge.}. \\

@@

\section{Chapter Summary}
In this chapter we investigated how the nonlinear solution deviates from the linear solution by including the effects of the nonlinearity. We did this numerically; First we had to validate the code by comparing to the analytic result from Chapter \ref{chap:linear}. Once the code yielded the expected results for the linearized version, we incorporated on the nonlinearity. \\

@@

To better understand the stability of the pulse, we turned our attention to computing the error between consecutive iterations by \eqref{eq:error}. This showed that the relative error within the wave breaking region is of order unity. Whereas, in the stable region the error took on values between $10^{-4}$ and $10^{-16}$. This is because the pulse reached a steady state with a periodicity greater than 1, this was highlighted by computing the composite error, \eqref{eq:deltaerror}, shown in Figure \ref{fig:errorcomp}. \\

>>>>

\subsection{Validation}
\begin{figure}[tbp]
\centering
\input{./Figures/Variance}
\caption[Equilibrium variance, chirp, and phase shift of the pulse as a function of the dispersion parameter.]{Simulation and analytic equilibrium variance, chirp, and phase shift as a function of the dispersion parameter, $s$. The analytic expressions are given by \eqref{eq:equilvar}, \eqref{eq:equilchirp}, \eqref{eq:phase}, respectively.}
\label{fig:var}
\end{figure}
Before delving into the nonlinear behaviour of the model, we wish to validate that the code is working as expected by comparing the results of the simulations with $b=0$ to the results of the linear model in Chapter \ref{chap:linear}. In the case of the numerical solution, our initial conditions become somewhat important for two reasons. First, if the energy is too high initially the nonlinear effects will be too strong leading to \acrlong{spm}. Secondly, the initial waveform needs to be well behaved enough so that the Fourier transform is band limited. In all of the following analysis the initial waveform is $\Gamma \sech \left( 2T \right) \textrm{e}^{i \pi / 4}$ normalized so that the initial energy is $E_0 = 0.1$; additionally, $a = 8000$, and $h = 0.04$. \\

@@

In the first experiment the pulse is allowed to equilibrate for 40 loops of the circuit\footnote{The choice of 40 will be justified in the following section.}, we then compare the variance, chirp, and phase shift of the analytic and numerical solutions. The results of this are shown in Figure \ref{fig:var}. For the most part, we see exceptional agreement between the analytic solution and numerical solution for the variance, and phase shift. However, as $s \rightarrow 2.5$ the chirp from the simulations begins to deviate from the analytic solution. The reason for this is quite a simple one. We shall turn our attention to the energy, and amplitude of the pulse at equilibrium as shown in Figure \ref{fig:valenergy}. As with Figure \ref{fig:var}, there is very good agreement between the two solutions. It is also now clear that at approximately $s = 2.5$ there is too much dispersion---the pulse is no longer sustainable. This is of course a consequence of the condition \eqref{eq:energycond}. Since the pulse effectively vanishes after this point, the chirp calculation becomes numerically unstable leading to the disagreement. \\

With the numerical solution yielding the expected results we are now ready to explore the rich structure that the nonlinearity adds.

\section{Nonlinear Model}
\begin{figure}[tbp]
\centering
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Shape}
\caption{Envelope}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_FT}
\caption{Fourier transform}
\end{subfigure} \\
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Chirp}
\caption{Chirp}
\end{subfigure}
\caption[Envelope, Fourier transform, and chirp of the pulse---stable case.]{Simulation with $s = 0.15$, and $b = 2.1$ after 15 circuits.}
\label{fig:nlstable}
\end{figure}
With the inclusion of the nonlinearity we generally find a similar solution to the linear case. An example of this is shown in Figure \ref{fig:nlstable}. The envelope of the pulse is unsurprisingly Gaussian-esque, however, it is \emph{not} precisely Gaussian and more closely resembles a generalized Gaussian\footnote{A generalized Gaussian has the form $\exp \left(-t^\alpha \right)$, with $\alpha > 2$.}. The fact that the pulse envelope is not Gaussian is further emphasized while examining the Fourier transform of the pulse. If the pulse were a Gaussian, we would expect the Fourier transform to also be a Gaussian \cite{debnath, gradshteyn}. Instead, the magnitude of the Fourier transform has a unique weakly bi-modal shape. This deviation suggests the nonlinearity implants higher frequency oscillations into the pulse---a key observation in the coming subsections. Finally, we shall examine the derivative of the phase---essentially the chirp. Recall in Chapter \ref{chap:linear} the chirp was defined as coefficient of $-\frac{1}{2}iT^2$ in the exponential, and by taking the negative of the derivative we would expect a linear function with a slope equal to the chirp. In the nonlinear case, we recover this linear response for moderate values of $T$. However, for $|T| > 1$ the chirp begins to saturate. This is consistent with the experimental results \cite{chen, rothenberg, tomlinson}. \\

@@

\begin{figure}[tbp]
\centering
\input{./Figures/ROC}
\caption{Rate of convergence as a function of number of iterations for various $s$ values, with $b = 0.5$. The error is given by \eqref{eq:error}.}
\label{fig:roc}
\end{figure}

@@

The energy of the pulse is directly related to the output power of the laser, and since this is not as controllable as in a regular laser, it is of great interest. In this nonlinear case we allow the initial pulse to pass through each component 100 times instead of 40 in the linearized case. This choice can be justified by Figure \ref{fig:roc}. The pulse reaches a state in which the error no longer decreases within typically 10--20 iterations. In the linearized case we chose to iterate 40 times since this is about double the number of iterations expected. Furthermore, in the nonlinear case we extend this number further to 100 in case the nonlinearity increases the number of iterations required to converge. The energy of the pulse after 100 iterations is shown in Figure \ref{fig:energy}. In Figure \ref{fig:energyzoom} we generally find that the energy is a smooth surface that slowly decays as $s$ and $b$ increase. In the upper left hand region of the plot this is clearly not the case. The contours show that the energy is very noisy with very large derivatives. Perhaps more surprising, as shown in Figure \ref{fig:energybig}, is that this boundary between periodic and aperiodic behaviour appears to become periodic (in the parameter space) as $b$ increases around $s = 0.3$. These features are further highlighted in Figure \ref{fig:cartoon} by disregarding the energy and only examining whether the pulse is invariant after additional iterations. \\

@@

The noise in the energy exhibited for moderate to large values of $b$, and small values of $s$ is a phenomenon called \emph{wave breaking} \cite{agrawal2013, anderson, finot, rothenberg, tomlinson}. Wave breaking is not limited to just optics; wave breaking occurs in areas such as plasmas, transmission lines, and fluid dynamics \cite{rothenberg}. Wave breaking occurs because the pulse begins to interfere with itself in a way called \gls{spm} \cite{agrawal2002, agrawal2013, becker}. \gls{spm} occurs because the index of refraction is intensity dependent \cite{agrawal2002, becker, rothenberg, silfvast}, which leads to additional chirp across the pulse \cite{agrawal2013, anderson, rothenberg, silfvast}. This in turn causes higher order frequencies to be injected into the pulse \cite{agrawal2013, anderson}, as we saw in Figure \ref{fig:nlstable}. These high frequencies compound with each trip around the cavity becoming parasitic very quickly---Figure \ref{fig:break} highlights this. Notice that the difference between Figure \ref{fig:nlstable} and Figure \ref{fig:break} is a difference in $b$ of $0.05$---just enough to cross the boundary---this difference could be attributed to adding a few centimetres more of fibre between the gain and output coupler. The left figures show the pulse after 11 trips around the cavity; in the Fourier transform it is clear that the contributions from higher frequencies has increased. We obtain comparable results as in the experiments \cite{anderson, rothenberg}. Additionally, the chirp starts losing its linearity causing it to start becoming unstable; the nature of this instability is again in agreement with the experiments \cite{anderson, rothenberg}. The parasitic nature of the high frequency contributions is evident by examining the right figures. After five additional trips around the cavity, the envelope of the pulse is much more rippled, and the real and imaginary parts become incoherent. Moreover, the Fourier transform has no clear structure and has essentially become noise. The chirp has grown to be highly oscillatory and unstable. Once the pulse has reached a state such as this, the envelope, Fourier transform, and chirp never reach a steady equilibrium state. \\

@@

Unsurprisingly, the error is largest in the region where the wave breaks. As mentioned in the previous subsection, the pulse does not reach a stable state in this region. As a consequence the envelope varies drastically, which leads to this large error. Conversely, the region where the energy appeared to be stable still has a non-zero error. The cause of this discrepancy is because some equilibrium states have a periodicity greater than $1$, thus, we redefine \eqref{eq:error} to be instead
\begin{align}
\textrm{E}_\Delta = \frac{\| |A_i| - |A_{i-\Delta}| \|_2}{\| A_{i-\Delta} \|_2},
\label{eq:deltaerror}
\end{align}
to pick up equilibrium states with period $\Delta$, again with $i = 100$. Figure \ref{fig:deltaerror} shows the error for an assortment of $\Delta$ values. There are several key observations to make, first, for $\Delta = 2$ the kidney-shaped blob in the lower right corner has vanished, thus, this entire region has a periodicity of $2$. Second, for $\Delta = 3$ in the upper right corner a negative triangular-shaped structure emerges. Lastly, $\Delta \in \left\{ 5, 7, 11, 13 \right\}$ may not seem to extract any higher order periods, however, there are in fact very small striations in the lower left. We typically chose $\Delta$ values to be prime since they are the building blocks for composite numbers. However, there are two composite $\Delta$ values of interest. The first is $16$, since $16 = 2^4$, as it will pick up any periodic behaviour with orders of $2, 4, 8, 16$. Lastly, we chose $\Delta = 60$ for a similar reason: because $60$ is highly divisible we can extract orders such as $12$ or $30$ without having to explicitly run the calculations\footnote{In the case of $\Delta = 60$ a value of $i = 150$ is used to ensure sufficient iterations to converge.}. \\

@@

\section{Chapter Summary}
In this chapter we investigated how the nonlinear solution deviates from the linear solution by including the effects of the nonlinearity. We did this numerically; first we had to validate the code by comparing to the analytic result from Chapter \ref{chap:linear}. Once the code yielded the expected results for the linearized version, we incorporated the nonlinearity. \\

@@

To better understand the stability of the pulse, we turned our attention to computing the error between consecutive iterations by \eqref{eq:error}. This showed that the relative error within the wave breaking region is of order unity. Whereas, in the stable region the error took on values between $10^{-4}$ and $10^{-16}$. This is because the pulse reached a steady state with a periodicity greater than 1, which was highlighted by computing the composite error, \eqref{eq:deltaerror}, shown in Figure \ref{fig:errorcomp}. \\
